Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-06 paper/paper2.tex
added final figures and got through muller results

\section{Background}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.

@@

The MSM models the \hl{action?} of the propagator with a Markov jump process on $\{1, \ldots, k\}$. Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one. The elements in $\mathbf{p}$ are defined by a coarse-graining of $p(\mathbf{x})$ over $S$, as
\begin{equation}
\label{eq:coarse-grain-p}
p_i = \int_{\mathbf{x} \in s_i} d\mathbf{x}\; p(\mathbf{x}).
\end{equation} 
Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
\begin{equation}
\mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
\end{equation}
where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$. With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$.

@@

@@

\section{Computational Methods}

@@

For the tICA MSMs, the cluster centers from the 100 state model were used as the test points $Y$ to define $A$ with a cutoff of $R=0.81$. In the PCA MSMs, the cluster centers from the 500 state model were used as the test points with a cutoff of $R=2.0$. In each case, $M=10^6$ rounds of Monte Carlo integration were performed to compute the state volumes.

For all models, a maximum likelihood estimator for reversible transition matrices at a given lag time, with a pseudocount of $1 / k$ was used to compute $\mathbf{T}$. The pseudocount ensures that all transitions are assigned nonzero probability, which is especially important for evaluating data that was not used to train the model, as often new transitions will be observed.

@@

\begin{figure}[h]
\centering
\includegraphics[width=4in]{figs_final/mull_likes.png}
\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

@@

 All MSMs were built using a lag time of 50 ns.

Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/ww_likes.png}
\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). \label{fig:ww}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs_final/timescales_suck2.pdf}
\caption{For MSMs of the WW domain built on the top three PCA components,} 
\end{figure}

>>>>

\section{Background}
Let $\mathbf{x}_t$ be a ergodic time-homogenous Markov process the phase space $\Omega$ with stationary density $\mu(\mathbf{x})$. Consider an ensemble of such processes at time $t$, described by a distribution $p_t(\mathbf{x})$. There exists an operator, termed the propagator, that evolves the a distribution forward in time at discrete intervals, $\tau$.
\begin{equation}
p_{t+\tau}(\mathbf{x}) = \mathcal{Q}^{(\tau)} \circ p_{t}(\mathbf{x})
\end{equation}
The eigenfunctions of the propagator form a basis for $p(\mathbf{x})$, and the time evolution of the ensemble after $n$ applications of the propagator can be decomposed into a sum of terms along each eigenfunction.
\begin{equation}
\label{eq:prop-timescales}
p_{t+n\tau}(\mathbf{x}) = \sum_{i=1}^\infty \exp\left(-\frac{n \cdot \tau}{t_i}\right) \langle \phi_i, p_t \rangle_{\mu^{-1}} \phi_i(\mathbf{x}),
\end{equation}
where $t_i = -\displaystyle\frac{\tau}{\ln \lambda_i}$ are the propagator's relaxation timescales and $\phi_i$ and $\lambda_i$ are the eigenfunctions and eigenvalues of the propagator, and $\langle \cdot, \cdot \rangle_{\mu^{-1}}$ denotes the inner product defined as
\begin{equation}
\langle \phi_i, p_t \rangle_{\mu^{-1}} = \int_\Omega d\mathbf{x}\; \phi_i(\mathbf{x}) p_t(\mathbf{x}) \frac{1}{\mu(\mathbf{x})}.
\end{equation}
For a more detailed discussion of the propagator and its properties, we refer the reader to \citet{Prinz2011Generation}.

@@

The MSM models the \hl{action?} of the propagator with a Markov jump process on $\{1, \ldots, k\}$. Let $\mathbf{p} \in (\mathbb{R} \cap [0,1])^k$ be a column vector whose entries sum to one. The elements in $\mathbf{p}$ are defined by a coarse-graining of $p(\mathbf{x})$ over $S$, as
\begin{equation}
\label{eq:coarse-grain-p}
p_i = \int_{\mathbf{x} \in s_i} d\mathbf{x}\; p(\mathbf{x}).
\end{equation} 
Consider a probability vector at time $t$ denoted $\mathbf{p}(t)$. The time-evolution of $\mathbf{p}(t)$ is described by
\begin{equation}
\mathbf{p}(t+\tau)^T = \mathbf{p}(t)^T \mathbf{T},
\end{equation}
where $T_{ij}$ is the probability of transiting to state $j$ in time $\tau$ given that the system started in state $i$. With this construction, the eigenvalues and eigenvectors of $\mathbf{T}$ \hl{provide approximations (what about transfer operator?)} for the eigenvalues and eigenfunctions of $\mathcal{Q}$. In a direct analogy with \cref{eq:prop-timescales}, the eigenvalues $\lambda_i$ of $T$ correspond to relaxation timescales of the Markov model
\begin{equation}
\label{eq:msm-timescales}
t_i = -\frac{\tau}{\ln \lambda_i}.
\end{equation}

@@

Selecting the size of $S$ is an important step in model construction, however there is not currently a widely-accepted method in use. Heuristic criteria, such as manually selecting $k = n_\textnormal{conformation} / 14$ or clustering $D$ such that each conformation is within a pre-specified distance to its cluster center (e.g. 4.5\AA$\;$ or 1.2\AA $\;$ RMSD) have been previously employed\cite{Lane2012, Bowman2012, Beuchamp}. Selection of $\tau$ based convergence of the relaxation timescales (\cref{eq:msm-timescales}), as suggested by \citet{Swope2004Describing} and others is more well established. The rate of this convergence with respect to $\tau$ at fixed $S$ has also been used as a metric for evaluating a model's state decomposition. \hl{cite someone?}

@@

\section{Methods}
\label{sec:methods}

@@

For the tICA MSMs, the cluster centers from the 100 state model were used as the test points $Y$ to define $A$ with a cutoff of $R=0.81$. In the PCA MSMs, the cluster centers from the 500 state model were used as the test points with a cutoff of $R=2.0$. In each case, $M=10^6$ rounds of Monte Carlo integration were performed to compute the state volumes. All WW MSMs were built using a lag time of 50 ns, which is the same lag time used in previous analyses\cite{Beauchamp2012Simple}.

For both the M\"{u}ller potential and WW models, a maximum likelihood estimator for reversible transition matrices, with a pseudocount of $1 / k$ was used to compute $\mathbf{T}$. The pseudocount ensures that all transitions are assigned nonzero probability, which is especially important for evaluating data that was not used to train the model, as often new transitions will be observed. All analysis and model construction was performed using MSMBuilder \cite{Beauchamp2011Msmbuilder2}.

@@

How well can likelihood-based methods select parameters for an MSM in the data-rich regime? To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see ``Methods"). We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence. The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states. The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small. For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend. Above $k=600$, the test log-likelihood stops increasing. These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data. On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics. As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.

Leaving out a portion of a molecular dynamics data set from the training to serve as a test set is costly proposition for the applications of MSMs to the study of complex system. For large bimolecular simulations characterized by long intrinsic timescales, sampling the potential exhaustively is a significant challenge\cite{Lane2012}, making it difficult to afford discarding half of the dataset during the fitting of an MSM. For this reason, we also computed two penalized likelihood model selection criteria, the AIC and BIC, which augment the training set log-likelihood with an explicit penalty on model complexity to avoid over fitting. Applied to our simulations of the M\"{u}ller potential, both the AIC and BIC penalize model complexity more strongly than the test set log-likelihood.

The likelihood-based methods are consistent with model selection based on maximization of the implied timescales. For this simple data set, the timescales are maximized by models with 200 states, which agrees with the results obtained with BIC. Model selection based on direct maximization of the implied timescales considers only the systematic error in the MSM, and neglects the statistical error. The consistency between the two approaches on this data set is an indication that the statistical uncertainty is low for these models, which is to be expected given the ease of sampling this two dimensional toy potential.

\begin{figure}[h]
\centering
\includegraphics[width=6in]{figs/mull/mull.pdf}
\caption{\hl{Fix this figure so that only part (A) is depicted.} We calculated the training and test set log-likelihoods and model selection criteria for models built with 50-1000 states using the uniform emission distribution. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). For both likelihoods, the BIC penalizes complexity more than the AIC and test set log-likelihood. The optimal window is consistent with the convergence of the relaxation timescales (right frames).}
\label{fig:mullerlike}
\end{figure}

@@

Because the likelihood calculations involving the uniform distribution emission model have exponential complexity in the dimensionality of the state space we first preprocess the trajectories with a dimensionality reduction technique.
\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}
\caption{\hl{Fix this figure so that only part (A) is depicted.} On models whose states were defined after projecting into the four-dimensional tICA subspace, we computed likelihoods based on the uniform emission models. Dashed lines represent the optimal model as defined by the AIC, BIC, or test set log-likelihood and form the boundaries of the optimal window (shaded dark gray). \label{fig:ww}}
\end{figure}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{For MSMs of the WW domain built on the top three PCA components,} 
\end{figure}
